Appendix 6: sdmTMB speed testing methods and model validation description

### 2022-05-06 doc/appendix-binomial-inla.Rmd
Double space; update appendices for latest sdmTMB

---
title: "Appendix 4: Dynamic species distribution model comparison between INLA, inlabru, and sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
---

>>>>

---
title: "Appendix 4: Dynamic species distribution model comparison between INLA, inlabru, and sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \linenumbers
---

### 2022-05-06 doc/appendix-pcod-VAST-tweedie.Rmd
Double space; update appendices for latest sdmTMB

---
title: "Appendix 5: Comparison of index standardization of fishery-independent survey data with VAST and sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: false
    number_sections: true
---

@@

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
pred <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE, area = 4)
ind_sdmTMB <- get_index(pred)
```

>>>>

---
title: "Appendix 5: Comparison of index standardization of fishery-independent survey data with VAST and sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: false
    number_sections: true
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \linenumbers
---

@@

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
pred <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE)
ind_sdmTMB <- get_index(pred, area = 4)
```

### 2022-05-06 doc/appendix-pcod-VAST.Rmd
Double space; update appendices for latest sdmTMB

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
tictoc::tic()
p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE,
  area = rep(4, nrow(qcs_grid))
)
ind <- get_index(p, bias_correct = FALSE)
tictoc::toc()
```

>>>>

```{r sdmTMB-index, cache=TRUE, echo=TRUE}
tictoc::tic()
p <- predict(fit_sdmTMB, newdata = qcs_grid, return_tmb_object = TRUE)
ind <- get_index(p, bias_correct = FALSE, area = rep(4, nrow(qcs_grid)))
tictoc::toc()
```

### 2022-05-06 doc/appendix-pcod-sdmTMB.Rmd
Double space; update appendices for latest sdmTMB

---
title: "Appendix 2: Example of index standardization of fishery-independent survey data with sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
bibliography: refs.bib
csl: mee.csl
---

@@

@@

```{r sim-resids, out.width="4in", warning=FALSE, results='hide'}
simulate(fit, nsim = 500) %>%
  dharma_residuals(fit)
```

@@

```{r sims}
pred_sims <- predict(fit, newdata = qcs_grid, sims = 500)
dim(pred_sims)
```

@@

```{r index, fig.asp=0.45, out.width="5in"}
qcs_grid$area <- 4 # all 2 x 2km
pred2 <- predict(
  fit,
  newdata = qcs_grid, return_tmb_object = TRUE,
  area = qcs_grid$area
)
ind <- get_index(pred2)
saveRDS(ind, "sdmTMB-index.rds")

@@

The above example used the default `bias_correct = FALSE` in `get_index()`. The `bias_correct` argument is a feature of TMB for models with random effects, and more details can be found in the [TMB documentation](https://github.com/kaskr/adcomp/wiki) or [@thorson2016bias]. Leaving `bias_correct = FALSE` is fine for experimentation or a quick example, but will be biased due to the nonlinear (`exp()`) transformation of the random effects. Setting `bias_correct = TRUE` implements a generic bias correction algorithm that should correct for this, but can take considerable time and memory.

As an alternative, we can use the output from the joint precision matrix simulation draws to calculate our index. This should approximate the output of `get_index(x, bias_correct = TRUE)`. However, we note that there may be situations where this approach may fail or be biased and experimentation is ongoing.

```{r index-sims, fig.asp=0.45, out.width="5in"}
ind <- get_index_sims(pred_sims, area = rep(4, nrow(pred_sims)), level = 0.95)

>>>>

---
title: "Appendix 2: Example of index standardization of fishery-independent survey data with sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
bibliography: refs.bib
csl: mee.csl
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \linenumbers
---

@@

An alternative would be to model catch as the response with an offset (`offset` argument) for log area swept.

@@

```{r sim-resids, out.width="4in", warning=FALSE, results='hide'}
set.seed(123)
simulate(fit, nsim = 500) %>%
  dharma_residuals(fit)
```

Or perhaps better, use MCMC residuals:

```{r mcmc-resids, warning=FALSE, results='hide', cache=TRUE}
set.seed(123)
r <- residuals(fit,
  type = "mle-mcmc", 
  mcmc_iter = 400, mcmc_warmup = 300
)
```

```{r mcmc-resids-vis, out.width="4in"}
qqnorm(r)
qqline(r)
```

@@

```{r sims}
pred_sims <- predict(fit, newdata = qcs_grid, nsim = 500)
dim(pred_sims)
```

@@

```{r index, fig.asp=0.45, out.width="5in"}
qcs_grid$area <- 4 # all 2 x 2km
pred2 <- predict(
  fit,
  newdata = qcs_grid, 
  return_tmb_object = TRUE
)
ind <- get_index(pred2, area = qcs_grid$area, bias_correct = TRUE)

@@

The above example uses `bias_correct = TRUE` in `get_index()`. The `bias_correct` argument is a feature of TMB for models with random effects, and more details can be found in the [TMB documentation](https://github.com/kaskr/adcomp/wiki) or [@thorson2016bias]. Leaving `bias_correct = FALSE` is fine for experimentation or a quick example, but will be biased due to the nonlinear (`exp()`) transformation of the random effects. Setting `bias_correct = TRUE` implements a generic bias correction algorithm that should correct for this, but takes more time and memory than skipping the bias correction.

### 2022-05-06 doc/appendix-snow-sdmTMB.Rmd
Double space; update appendices for latest sdmTMB

---
title: "Appendix 3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
urlcolor: blue
bibliography: refs.bib
csl: mee.csl
---

@@

## Basic inspection
```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=TRUE}
if (!rds_saved) {
  saveRDS(m0, here::here("data/snow_w_main_effect_0_150km.rds"))
  saveRDS(m1, here::here("data/snow_w_main_effect_1_150km.rds"))
  saveRDS(m2, here::here("data/snow_w_main_effect_150km.rds"))
}
```

@@

According to AIC, the `nbinom2()` (negative binomial, NB2) distribution appears to have the most parsimonious fit. 
We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and check that the gradients are all small enough.
If the gradients aren't quite small enough, we can easily run additional optimization steps using the `run_extra_optimization()` function.

@@

```{r fit-negb2-reml, echo=TRUE, eval=!rds_saved2, warning=FALSE, cache=FALSE}
m <- sdmTMB(count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ 0 + nao,
  time = "year", 
  spatial = "on", 
  spatiotemporal = "iid",
  family = nbinom2(link = "log"),
  reml = TRUE,
  data = snow, mesh = mesh,
  silent = TRUE
)

if (max(m$gradients) > 0.01) {
  m <- run_extra_optimization(m, newton_loops = 1L, nlminb_loops = 0L)
}
```

@@

```{r predict, message=FALSE}
pred <- predict(m)
pred$resid <- residuals(m)
```

@@

More accurate residuals can be generated using tmbstan, but this is a very slow process!
In this case, it took more than 2.5 hrs.

```{r stan-resids, eval=FALSE}
stan_fit <- tmbstan::tmbstan(m$tmb_obj, iter = 100, chains = 1, warmup = 99)
eta <- predict(m, tmbstan_model = stan_fit)
mu <- m$family$linkinv(as.numeric(eta))
r_mcmc <- sdmTMB:::qres_nbinom2(m, y = m$data$count, mu = mu)
saveRDS(r_mcmc, "data/mcmc_residuals.rds")
```

```{r stan-dat, echo=FALSE, cache=FALSE}
r_mcmc <- readRDS(here::here("data/mcmc_residuals.rds"))
```

```{r stan-qqnorm, out.width="33%"}
qqnorm(r_mcmc)
qqline(r_mcmc)
```

We can also test for zero-inflation using the `simulate.sdmTMB()` function and a couple functions from the DHARMa package.
If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead. 
This is not built directly into sdmTMB, but can be achieved by fitting the two components separately. See the steps in the vignette [Fitting delta models with sdmTMB](https://pbs-assess.github.io/sdmTMB/articles/delta-models.html).

@@

<!-- ## Cross-validation -->

<!-- Cross-validation can also be used to assess model performance. -->
<!-- Here, we use the `future` package to speed up the running of multiple models and save a smaller version of the output that doesn't include the full model information. -->
<!-- We use `k_folds = 5`, which amounts to training on 80% of the data and testing with 20% for each model. -->
<!-- Block cross-validation, which can be done using custom folds defined using the `fold_ids` argument, is generally recommended for spatiotemporal data. -->
<!-- However, random folds created by `sdmTMB_cv()` are appropriate in this case, because we are only interested in predicting within the spatial and temporal range of our data and are attempting to estimate the spatial variation in the effect of NAO. -->

<!-- ```{r cv, eval=FALSE, cache=FALSE} -->
<!-- library(future) -->
<!-- plan(multisession) -->
<!-- options(future.rng.onMisuse = "ignore") -->

<!-- m_cv <- sdmTMB_cv(count ~ 1 + nao + (1 | year_f), -->
<!--   spatial_varying = ~ 0 + nao, -->
<!--   time = "year", -->
<!--   spatial = "on", -->
<!--   spatiotemporal = "iid", -->
<!--   family = nbinom2(link = "log"), -->
<!--   reml = TRUE, -->
<!--   use_initial_fit = TRUE, -->
<!--   data = snow, mesh = mesh, -->
<!--   k_folds = 5 -->
<!-- ) -->
<!-- plan(sequential) -->
<!-- saveRDS(m_cv, file = here::here("data/owlcvallreml.rds")) -->
<!-- # m_cv$models[[1]] # to retrieve just the model built excluding the first fold -->
<!-- cv <- m_cv -->
<!-- cv$models <- NA -->
<!-- saveRDS(cv, file = here::here("data/owlcvdatreml.rds")) -->
<!-- ``` -->

<!-- First, we should check that all the models converged and have gradients < 0.01. -->

<!-- ```{r explore-cv, message=FALSE, cache=FALSE} -->
<!-- cv <- readRDS(here::here("data/owlcvdatreml.rds")) -->
<!-- cv$converged -->
<!-- max(cv$max_gradients) -->
<!-- ``` -->

<!-- If we were conducting model selection, we might contrast the log likelihoods (`cv$sum_loglik`), and the expected log predictive density (`cv$elpd`) for the left out data between different model configurations instead of relying on AIC. -->
<!-- We can also check the prediction accuracy on left-out data by plotting the `cv_predicted` against the observed `count` in log space. -->

<!-- ```{r cv-pred-obs, out.width="70%", cache=FALSE} -->
<!-- cvdat <- cv$data %>% filter(cv_loglik > -60) # this removes one extreme outlier -->
<!-- ggplot(cvdat, aes(log(count + 1), log(cv_predicted + 1), colour = -cv_loglik)) + -->
<!--   geom_abline(slope = 1, colour = "gray") + -->
<!--   geom_point(size = 2, alpha = 0.25) + -->
<!--   scale_colour_viridis_c( -->
<!--     direction = -1, option = "A", begin = 0.2, end = 0.9 -->
<!--   ) + -->
<!--   coord_fixed(ylim = c(0, 3.8)) + -->
<!--   theme_bw() -->
<!-- ``` -->

<!-- These predictions for left-out locations match the observed counts just as well as those based on the model built using all the data shown below. -->
<!-- # Does this mean that we are at least not over-fit, despite the spatiotemporal random fields and spatially varying coefficient? -->

<!-- ```{r full-pred-obs, echo=FALSE, out.width="70%"} -->
<!-- ggplot(pred, aes(log(count + 1), log(exp(est) + 1), colour = resid)) + -->
<!--   geom_abline(slope = 1, colour = "gray") + -->
<!--   # geom_point(size = 0.75, alpha=0.5) + facet_wrap(~year) + -->
<!--   geom_point(size = 2, alpha = 0.25) + -->
<!--   coord_fixed(ylim = c(0, 3.8)) + -->
<!--   scale_colour_viridis_c( -->
<!--     "residuals", -->
<!--     direction = -1, option = "A", begin = 0.2, end = 0.9 -->
<!--   ) + -->
<!--   theme_bw() #+ theme(strip.background = element_rect(fill = NA, colour = NA)) -->
<!-- ``` -->

<!-- We can also explore the spatiotemporal variation in the predictive accuracy of our model by plotting the `cv_loglik`. -->

<!-- ```{r cv-map, cache = FALSE} -->
<!-- ggplot(cvdat, aes(X, Y, colour = -cv_loglik)) + -->
<!--   geom_point(size = 0.25) + -->
<!--   scale_colour_viridis_c( -->
<!--     direction = -1, option = "A", begin = 0.2, end = 0.9, -->
<!--     guide = guide_colourbar(title.position = "top") -->
<!--   ) + -->
<!--   coord_fixed() + -->
<!--   theme(axis.text = element_blank(), axis.ticks = element_blank()) + -->
<!--   facet_wrap(~year) -->
<!-- ``` -->

@@

nsim <- 500
zeta_sim <- predict(m, nsim = nsim, sims_var = "zeta_s")
sims <- spread_sims(m, nsim = nsim)
beta <- sims$nao
combined <- beta + t(zeta_sim)

p$zeta_sim <- apply(t(zeta_sim), 2, median)
p$nao_effect_sim <- apply(combined, 2, median)
p$nao_effect_lwr <- apply(combined, 2, quantile, probs = 0.10)
p$nao_effect_upr <- apply(combined, 2, quantile, probs = 0.90)

p_mean <- p %>%
  group_by(CBCID) %>% summarise(
    year = max(year),
    X = mean(X), Y = mean(Y),
    mean_est_count = exp(mean(est)),
    nao_effect = b[b$term == "nao", 2] + mean(zeta_s),
    nao_sim = mean(nao_effect_sim),
    nao_lwr = mean(nao_effect_lwr),
    nao_upr = mean(nao_effect_upr)
  )
```

```{r}
nao_effect_lwr95 <- apply(combined, 2, quantile, probs = 0.025)
```

>>>>

---
title: "Appendix 3: Example of a spatially varying effect of climate on citizen science count data with sdmTMB"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
urlcolor: blue
bibliography: refs.bib
csl: mee.csl
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \linenumbers
---

@@

## Basic inspection

```{r save-models, warning=FALSE, message=FALSE, echo=FALSE, cache=FALSE, eval=TRUE}
if (!rds_saved) {
  saveRDS(m0, here::here("data/snow_w_main_effect_0_150km.rds"))
  saveRDS(m1, here::here("data/snow_w_main_effect_1_150km.rds"))
  saveRDS(m2, here::here("data/snow_w_main_effect_150km.rds"))
}
```

@@

According to AIC, the `nbinom2()` (negative binomial, NB2) distribution has the most parsimonious fit.
We could also check residual distributions at this stage.
We will rerun with `reml = TRUE` to ensure better estimation of the random effects that dominate this model and check that the gradients are all small enough.
If the gradients aren't quite small enough, we can easily run additional optimization steps using the `run_extra_optimization()` function.

```{r sanity, echo=TRUE}
sanity(m2)
```

@@

```{r fit-negb2-reml, echo=TRUE, eval=!rds_saved2, warning=FALSE, cache=FALSE}
m <- sdmTMB(count ~ 1 + nao + (1 | year_f),
  spatial_varying = ~ 0 + nao,
  time = "year", 
  spatial = "on", 
  spatiotemporal = "iid",
  family = nbinom2(link = "log"),
  reml = TRUE,
  data = snow, mesh = mesh,
  silent = TRUE
)
```

@@

```{r predict, message=FALSE}
pred <- predict(m)
set.seed(19208)
pred$resid <- residuals(m)
```

@@

More accurate residuals can be generated using tmbstan, but this a slower process.
These are randomized quantile residuals where the fixed effects are fixed at their MLE (maximum likelihoood estimate) values and the random effects are sampled with MCMC via tmbstan/Stan.
To calculate these, we have to use the version of our model fit without REML.
We would ideally sample with more MCMC samples than this and assess convergence by setting `print_stan_model = TRUE`.

```{r stan-resids}
f <- here::here("data/mcmc_residuals.rds") # for caching
set.seed(82919)
if (!file.exists(f)) {
  mcmc_resids <- residuals(m2, type = "mle-mcmc", mcmc_iter = 101, mcmc_warmup = 100)
  saveRDS(mcmc_resids, f)
} else {
  mcmc_resids <- readRDS(f)
}
```

```{r stan-qqnorm, out.width="33%"}
qqnorm(mcmc_resids)
qqline(mcmc_resids)
```

We can also test for zero-inflation using the `simulate.sdmTMB()` function and a couple functions from the DHARMa package.
If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead (e.g., `delta_truncated_nbinom2()` or possibly even `delta_gamma()` if we were willing to treat the counts as continuous.

@@

@@

nsim <- 500
zeta_s <- predict(m, nsim = nsim, sims_var = "zeta_s")
sims <- spread_sims(m, nsim = nsim)
beta <- sims$nao
combined <- beta + t(zeta_s)

p$zeta_s <- as.numeric(apply(t(zeta_s), 2, median))
p$nao_effect_sim <- as.numeric(apply(combined, 2, median))
p$nao_effect_lwr <- as.numeric(apply(combined, 2, quantile, probs = 0.10))
p$nao_effect_upr <- as.numeric(apply(combined, 2, quantile, probs = 0.90))

p_mean <- p %>%
  group_by(CBCID) %>% summarise(
    year = max(year),
    X = mean(X), Y = mean(Y),
    mean_est_count = exp(mean(est)),
    nao_effect = b[b$term == "nao", "estimate"] + mean(zeta_s),
    nao_sim = mean(nao_effect_sim),
    nao_lwr = mean(nao_effect_lwr),
    nao_upr = mean(nao_effect_upr)
  )
```

```{r}
nao_effect_lwr95 <- as.numeric(apply(combined, 2, quantile, probs = 0.025))
```

### 2022-05-06 doc/appendix-validation.Rmd
Double space; update appendices for latest sdmTMB

---
title: "Appendix 6: sdmTMB speed testing methods and model validation description"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
number_sections: false
fig_caption: true
highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
---

@@

We conducted 20 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values.

@@

This appendix illustrates a comparison of penalized smoothers with mgcv and a comparison of IID random intercepts with glmmTMB.
<!-- Additionally, it illustrates some simulation testing of a spatial random field model with the main observation families within the sdmTMB package. -->

@@

```{r plot-smooth, echo=TRUE, fig.asp=0.6, out.width="5in"}
plot_smooth(m)

x2 <- seq(0, 1, length.out = 100)
p_mgcv <- predict(m_mgcv, newdata = data.frame(x2 = x2), se.fit = TRUE)
qv <- stats::qnorm(1 - (1 - 0.95) / 2)
df <- data.frame(
  x2 = x2,
  est = p_mgcv$fit, 
  lwr = p_mgcv$fit - qv * p_mgcv$se.fit, 
  upr = p_mgcv$fit + qv * p_mgcv$se.fit
)
lines(df$x2, df$est + 0.02, col = "blue")
lines(df$x2, df$lwr + 0.02, lty = 2, col = "blue")
lines(df$x2, df$upr + 0.02, lty = 2, col = "blue")
```

>>>>

---
title: "Appendix 6: sdmTMB speed testing methods and model validation description"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
number_sections: false
fig_caption: true
highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \linenumbers
---

@@

We conducted 50 simulation iterations per model and mesh resolution and show results for the mean, lower, and upper values.
The models were fit on a 2022 14 inch M1 MacBook Pro with an Apple M1 Pro 10-core CPU and 32 GB of RAM in R 4.1.3 and the default Matrix library. The following package versions were used:

```{r pkg-versions}
packageVersion("INLA")
packageVersion("inlabru")
packageVersion("mgcv")
packageVersion("spaMM")
packageVersion("sdmTMB")
```

@@

This appendix illustrates a comparison of penalized smoothers with mgcv and a comparison of IID random intercepts with glmmTMB.

@@

```{r plot-smooth, echo=TRUE, fig.asp=0.8, out.width="5in"}
par(cex = 0.8)
# sdmTMB:
plot_smooth(m)
# also see the more general visreg::visreg()

# mgcv:
x2 <- seq(0, 1, length.out = 100)
p_mgcv <- predict(m_mgcv, newdata = data.frame(x2 = x2), se.fit = TRUE)
qv <- stats::qnorm(1 - (1 - 0.95) / 2)
df <- data.frame(
  x2 = x2,
  est = p_mgcv$fit, 
  lwr = p_mgcv$fit - qv * p_mgcv$se.fit, 
  upr = p_mgcv$fit + qv * p_mgcv$se.fit
)
lines(df$x2, df$est + 0.02, col = "blue")
lines(df$x2, df$lwr + 0.02, lty = 2, col = "blue")
lines(df$x2, df$upr + 0.02, lty = 2, col = "blue")
```
